---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(tmap)
library(knitr)
library(sjPlot)
library(car)
library(lmtest)

```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}

hist(lab6_data$ndvi_20_med)

hist(lab6_data$maxtemp_20_med)

hist(lab6_data$mintemp_20_med)

hist(lab6_data$rain_20_sum)

hist(lab6_data$pop_20)

hist(lab6_data$water_20_pct)

hist(lab6_data$elev_med)

```

{The NDVI and rainfall variables are skewed to the right. Whereas, the maximum and minimum temperature values are skewed to the left. The total population and percentage of land covered by water have a similar distribution with both variables heavily skewed to the right. The variable closest to normality or a bell shape is median elevation. However, this variable is still skewed to the right.}

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}

map1<-tm_shape(lab6_data) +
  tm_polygons("ndvi_20_med",style="jenks")

map2<-tm_shape(lab6_data) +
  tm_polygons("maxtemp_20_med",style="jenks")

map3<-tm_shape(lab6_data) +
  tm_polygons("mintemp_20_med",style="jenks")

map4<-tm_shape(lab6_data) +
  tm_polygons("rain_20_sum",style="jenks")

map5<-tm_shape(lab6_data) +
  tm_polygons("pop_20",style="jenks")

map6<-tm_shape(lab6_data) +
  tm_polygons("water_20_pct",style="jenks")

map7<-tm_shape(lab6_data) +
  tm_polygons("elev_med",style="jenks")

tmap_arrange(map1,map2,map3,map4,map5,map6,map7)

```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

{In 2020, the higher concentration of vegetation for NDVI (0.36 and greater) was located all along the edges and coastal areas (northern, eastern, and southwest areas) of the country. For the median maximum temperature variable, the higher temperatures (33.75 degrees celsius and greater) are located in the northern half of Australia with median minimum temperature (19.9 degrees celsius and greater) showing a similar pattern. Higher sums of rainfall were located along the western and northern coasts. Interestingly, the highest concentration of population (172,427 people and greater) was found on the west coast. The highest percentage of land (10.8% and greater) covered by water were are located in the central eastern and southwestern areas. Median elevation was higher along the east coast and western half of the country (484 meters and greater), with the highest elevation found in the central western area. There seems to be an overlap in high amounts and concentration between NDVI and the population, rain, water (southwest corner), and elevation (southwest corner) variables.}

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
model_maxtemp<-lm(ndvi_20_med~maxtemp_20_med,data=lab6_data)
summary(model_maxtemp)

model_mintemp<-lm(ndvi_20_med~mintemp_20_med,data=lab6_data)
summary(model_mintemp)

model_rain<-lm(ndvi_20_med~rain_20_sum,data=lab6_data)
summary(model_rain)

model_pop<-lm(ndvi_20_med~pop_20,data=lab6_data)
summary(model_pop)

model_water<-lm(ndvi_20_med~water_20_pct,data=lab6_data)
summary(model_water)

model_elev<-lm(ndvi_20_med~elev_med,data=lab6_data)
summary(model_elev)

```

{ The median maximum temperature and median minimum temperature variables exhibited similar summary results. Each variable's coefficient had a negative relationship, weak magnitude, and had a highly significant p-value. The median max temp variable had a coefficient of -0.130902, while the median min temp coefficient was -0.12282. However, the total population, rain, and elevation variables displayed positive relationship, weak magnitude, and were highly significant. The coefficient for the total population variables was 1.500e-06, and the coefficient for total rain was 9.124e-07.The median max temp, min temp, total pop, and rain variables had the same p-value of 2e-16. The elevation variable had a coefficient of 1.787e-04 and a p-value of 1.14e-09. Different from the other variables, the percentage of land covered by water variable had a insignificant p-value (0.249). The coefficient for this variable was -0.178263, showing a negative relationship and a weak magnitude }

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}

model_all<-lm(ndvi_20_med~maxtemp_20_med+rain_20_sum+pop_20+elev_med,data=lab6_data)
summary(model_all)

```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

{The coefficient for median max temp for this model is -1.170e-02, which shows a smaller magnitude compared to the previous model (-0.130902). A similar trend can be detected for the rain variable with a coefficient of 8.469e-07 compared to the previous 9.124e-07. The p-values stayed the same for both variables (2e-16) with is highly significant. The pop variable also showed a decrease in magnitude in this model with a coefficient of 2.862e-07 compared to 1.500e-06 in the previous model. Additionally, the p-value (0.00627) was not as significant. Additionally, the elevation variable showed a slight decrease in magnitude for the coefficient (1.233e-04) but had a slighlty more significant p-value 1.46e-11. 

The R2 value of 0.639 tells me the strength of this model is moderate. The p-value (2.2e-16) of the F-statistic suggest this model is significant and that one or more of the independent variables is a predictor of the dependent variable of NDVI.}

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}

lab6_data$residuals<-residuals(model_all)

hist(lab6_data$residuals)

tm_shape(lab6_data) +
  tm_dots("residuals",size=.2)

```

{The histogram shows the residuals are left or negatively skewed. There are outliers within a negative number, while there also be a higher frequency of residuals (over 200 range) around 0.00. The larger residuals are found along the coast of Austrailia, specifically the southern and Northern coasts. These values mainly reflect negative residuals.}

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
vif(model_all)

bptest(model_all)

```

{For the multicollinearity test, the values for all four independent variables was under 3, ranging from 1.009305 to 1.173863. These values imply there may be some correlation among the variables. Based on the p-value (2.2e-16) of the Breusch-Pagan test,there is likley heteroskedasticity in the model. We can reject the null hypothesis that the data is uniform.}

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

{Based on this model, median maximum temperature, median elevation, total population, and sum of population are important variables to consider in predicting NDVI. While the p-values in the summary table were significant, there was presence of heteroskedastity within the model. In order to get more significant results and stronger magnitude for these variable coefficient, it may be necessary to change one or more of the variables.}


**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

{I used Google and ChatGPT to revisit the meaning of the different values and items found in regression summary models. I found these tools to be helpful in my interpretation of the labd results. I also used these same materials to revisit the purpose of multicollinearity or heteroskedastity tests.}

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

{I feel good about the work I did for this lab and rate this lab as moderate. The biggest things I learned by completing the lab was interpreting the summary results for regression models, as well as multicollinearity or heteroskedastity tests.}


**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}

```




#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}

```

